\section{Matter (\textit{Rūpa})}

Welcome to the fifth lesson of this Practical Abhidhamma Course. This lesson describes Matter, the third of the four Ultimate Realities: \textit{citta}, \textit{cetasika}, \textit{rūpa}, and \textit{Nibbāna}.\footnote{See Chapter 6 of “A Comprehensive Manual of Abhidhamma” and “The Buddhist Teaching on Physical Phenomena (see Footnote 2 for links).}

The Pāḷi word “\textit{rūpa}” is usually translated as matter, but as we will see, \textit{rūpa} includes more than what is conventionally called matter; \textit{rūpa} includes all non-mental phenomena. During a talk on \textit{rūpa}, I heard a monk joke, “What is matter? Never mind. What is mind? Doesn’t matter.”\footnote{Believe it or not, this is the first joke in the first Simpsons cartoon (in 1987, Simpsons cartoons were two minute segments, before they became a series). In the opening scene of the cartoon, Bart asks Homer, ``What is the mind?": \url{https://www.youtube.com/watch?v=F988JDDuFsQ}} The mind is very important because spiritual development is the training of the mind. \textit{Rūpa} is important because it is an object of the mind, it plays a role in bringing sense objects into the mind, and the physical body is made up of \textit{rūpas}.

Consciousness, Mental Factors and \textit{rūpas} are impermanent; they arise for a moment and then pass away. It may seem that \textit{rūpas} such as \textbf{Hardness} or \textbf{Heat} can last, but the \textbf{Hardness} or the \textbf{Heat} that is experienced now is different from the \textbf{Hardness} or the \textbf{Heat} that arose a moment ago.

When analyzing a cup of coffee, a scientist might break it into solid (the cup), liquid (the coffee), and gas (the aroma). According to the Abhidhamma, a cup of coffee consists of \textit{rūpas} such as \textbf{Hardness} and \textbf{Heat}. The Abhidhamma focus is on how the cup of coffee impacts the senses. The focus of Buddhism is spiritual development which involves the mind, so the Abhidhamma approach uses the mind as the starting point. This is why in the example of the cup of coffee, the Abhidhamma focuses on how the object impacts the senses. The Buddha stressed the importance of viewing things as being made up of \textit{rūpas}, saying this perspective supported spiritual development.\footnote{MN 33: \url{http://www.accesstoinsight.org/tipitaka/mn/mn.033.than.html\#dhatu4}}

In the past two lessons, I shared a \textbf{\textit{RADICAL}} practice: \textbf{\textit{R}} is for Recognize, \textbf{\textit{A}} is for Accept, \textbf{\textit{D}} is for Depersonalize, \textbf{\textit{I}} is for Investigate, \textbf{\textit{CA}} is for Contemplate \textit{Anicca} or Contemplate \textit{Anattā} and \textit{\textbf{L}} is for Let go. Previously, I discussed the \textbf{\textit{RADICAL}} practice using Mind Moments and Mental Factors, but the same \textbf{\textit{RADICAL}} practice can be used with \textit{rūpas}.

If I want to drive from one place to another, I will consult a map. The map will show me the road but the map makes no reference to countless details such as the types of people, animals or trees that I might encounter. These details are irrelevant to the purpose of the map. If I want to develop myself spiritually, I will consult the Buddha’s teachings. The Dhamma will give practical advice on how to develop myself spiritually but makes no reference to unrelated topics such as physics and chemistry. These details are irrelevant to the purpose of the Buddha’s teachings.

Traditional science is purely objective with no importance placed on the role of the observer. However, modern physics, particularly quantum mechanics, has introduced the role of the observer. One of the founders of quantum mechanics wrote, “The path of a particle only comes into existence once it is observed.”\footnote{Werner Heisenberg: \url{http://www.aip.org/history/heisenberg/p08c.htm}} Buddhism places the observer at the absolute centre. From the Buddhist perspective, all external things are equivalent; they are relevant because of their impact on the senses of the observer. This approach makes sense because the focus of Buddhism is spiritual development and this means guarding the senses of the observer.

\subsection*{Four Great Elements}

\begin{figure} [H]

\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}

\noindent\begin{tabular}{P{0.28\textwidth} | C{.185\textwidth} C{0.165\textwidth} C{0.175\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Hardness}/\textbf{Softness}\newline \textbf{Roughness}/\textbf{Smoothness}\newline \textbf{Heaviness}/\textbf{Lightness}\newline \textit{Paṭhavī} & Hardness & Acting as a foundation & Receiving & The other three Great Elements &\\[15mm]
\textbf{Flowing}/\newline \textbf{Cohesion}\newline \textit{Āpo} & Trickling/\newline Oozing & Intensifying & Holding together & The other three Great Elements &\\[15mm]
\textbf{Heat}/\newline \textbf{Coldness}\newline \textit{Tejo} & Heat & Maturing/\newline Maintaining & Continued supply of softness & The other three Great Elements &\\[15mm]
\textbf{Supporting}/\newline \textbf{Pushing}/\newline \textbf{Pressure}\newline \textit{Vāyo} & Distending & Causing motion & Conveying & The other three Great Elements &\\[15mm]
\bottomrule
\end{tabular} 

\caption[]{The four Great Elements. The literal meanings of \textit{paṭhavī}, \textit{āpo}, \textit{tejo} \& \textit{vāyo} are earth, water, fire \& air.\footnotemark}

\end{figure}

\footnotetext{Defined in Visuddhimagga XI.93, explained in Chapter 1 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for links).}

In the Suttas, the Buddha defined \textit{rūpas} as consisting of the Four Great Elements and \textit{rūpas} that depend on these elements.\footnote{MN 28: \url{http://www.accesstoinsight.org/tipitaka/mn/mn.028.than.html}\\ The four Great Elements support each other through conascence (\textit{sahajāta}), conascence support (\textit{sahajātanissaya}), conascence presence (\textit{sahajātātthi}), mutuality (\textit{\textit{aññamañña}}), and non-disappearance (\textit{avigata}) conditions. The four Great Elements support the other \textit{rūpas} through the same set of conditions, excluding mutuality (\textit{\textit{aññamañña}}) condition. See Chapters 6, 7 and 17 of “The Conditionality of Life” (see Footnote 2 for link).} The Four Great Elements are earth-element, water-element, fire-element, and air-element.\footnote{The Pāḷi terms are \textit{paṭhavī-dhātu} (earth-element), \textit{āpo-dhātu} (water-element), \textit{tejo-dhātu} (fire-element), and \textit{vāyo-dhātu} (air-element).} Nowhere in the Suttas does the Buddha list the \textit{rūpas} that depend on the Four Great Elements. The \textit{rūpas} that depend on the Four Great Elements are defined in the Abhidhamma and in the Commentaries.

The word “element” tells us to focus on the properties or the qualities rather than on the literal meaning of earth, water, fire or air. The Four Great Elements are experienced, Ultimate Realities, not concepts. The ancient Greeks and others\footnote{\url{http://en.wikipedia.org/wiki/Classical_element}} used the same four names (earth, water, fire and air) to describe the composition of the world, but generally adopted the literal or symbolic meaning rather than the Buddhist approach that focuses on properties or qualities. \textit{Rūpas} are not abstract categories; they are the realities that appear in daily life.

\pagebreak

Earth-element is the \textbf{Hardness} or \textbf{Softness}, \textbf{Roughness} or \textbf{Smoothness}, \textbf{Heaviness} or \textbf{Lightness}, that is experienced. While walking, one experiences the \textbf{Hardness} of the floor against the soles of the feet. You can be aware of \textbf{Hardness} in the teeth; bite them together and feel how hard they are. You can be aware of \textbf{Softness} by pressing your tongue against the inside of your lips to feel its \textbf{Softness}. Rub your tongue over the edge of your teeth or brush your hand over the skin of your arm and feel \textbf{Roughness}. You can experience \textbf{Smoothness} by wetting your lips and rubbing your tongue over them, from side to side. Place one hand on top of the other in your lap and feel the \textbf{Heaviness} of the top hand, or feel the \textbf{Heaviness} of the head by bending it forward. Look for \textbf{Lightness} by wagging a finger up and down and feeling its lightness. If something floats in the ocean, it is because the earth-element in the ocean supports it. If a balloon floats, it is because the earth-element in the air supports it.

Water-element is \textbf{Flowing} and \textbf{Cohesion}. \textbf{Flowing} and \textbf{Cohesion} cannot be experienced by the senses, but can be understood by the mind. While walking, one experiences \textbf{Hardness}, \textbf{Heat} and \textbf{Pushing} arise together as one, and this is \textbf{Cohesion}. Imagine the blood \textbf{Flowing} through the blood vessels, the air \textbf{Flowing} into the lungs, or heat \textbf{Flowing} throughout the body. To understand \textbf{Cohesion}, consider how the body is held together by the skin, flesh, and sinews. The blood is held inside by the skin, like water in a balloon. Without \textbf{Cohesion}, the body would fall into separate pieces. The force of gravity which keeps the body stuck to the earth is also \textbf{Cohesion}.

Fire-element is the \textbf{Heat} and \textbf{Coldness} that is experienced. While walking, one experiences the \textbf{Coldness} of the floor against the soles of the feet. It is usually very easy to look for \textbf{Heat} throughout the body. Feel the \textbf{Coldness} of the breath as it enters the nostrils and then understand it throughout the body. The function of fire-element is to ripen, mature, age and burn up. The fire-element also aids in digestion.

Air-element is the \textbf{Supporting}, \textbf{Pushing}, and \textbf{Pressure} that is experienced. When walking, the movement of the feet includes an experience of \textbf{Pushing}. To experience \textbf{Supporting}, relax your back so your body bends forward. Then straighten it and keep it straight. The force that keeps the body straight is \textbf{Supporting}. The \textbf{Supporting} air-element ensures that all things maintain their shape and do not collapse. To experience \textbf{Pushing}, be aware through the sense of touch of \textbf{Pushing} in the centre of the head as you breathe in and out, or experience the \textbf{Pushing} as the chest expands or as the abdomen moves when breathing. Wherever there is movement, there is \textbf{Pushing}. Though the function of air-element is causing motion, movement is not the alteration between two stages of the same group of \textit{rūpas}, but rather the disappearance of one group of \textit{rūpas}, and the immediate emergence in its place of another group of \textit{rūpas}. For example, in a movie theatre, many static frames are projected onto the screen each second and a sense of movement is created.

All groups of \textit{rūpas} include the Four Great Elements. In a stone and a snowflake, the Four Great Elements exist in equal proportion, but of varying intensity. The intensity and the Mental Factor of \textbf{Attention} determine which experience body-consciousness recognizes first. For example, when you put your hand in water, the \textbf{Softness} felt is earth-element, the \textbf{Coolness} felt is fire-element and the \textbf{Pressure} felt is air-element. If the water is cold, then the fire-element may be experienced first. If attention is paid to the sensation of \textbf{Pressure}, then air-element may be experienced first.

\subsection*{Sense Objects}

\begin{figure} [H]
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}
\noindent\begin{tabular}{P{0.25\textwidth} | C{.185\textwidth} C{0.185\textwidth} C{0.185\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Visible-form}\newline \textit{Rūpa} & Striking the\newline eye & Object of\newline seeing process & Being the field of seeing process & Four Great Elements &\\[9mm]
\textbf{Sound}\newline \textit{Sadda} & Striking the\newline ear & Object of\newline hearing process & Being the field of hearing process & Four Great Elements &\\[9mm]
\textbf{Odour}\newline \textit{Gandha} & Striking the\newline nose & Object of\newline smelling process & Being the field of smelling process & Four Great Elements &\\[9mm]
\textbf{Taste}\newline \textit{Rasa} & Striking the tongue & Object of\newline tasting process & Being the field of tasting process & Four Great Elements &\\[9mm]

\bottomrule
\end{tabular}
\caption[]{The four sense objects.\footnotemark}

\end{figure}

\footnotetext{Explained in Chapter 4 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link).}

Earth-element, fire-element, and air-element are objects of the tactile sense, the sense of touch. The next set of \textit{rūpas} (\textbf{Visible-form},\footnote{Defined in Visuddhimagga XIV.54 (see Footnote 2 for link).} \textbf{Sound},\footnote{Defined in Visuddhimagga XIV.55 (see Footnote 2 for link).} \textbf{Odour},\footnote{Defined in Visuddhimagga XIV.56 (see Footnote 2 for link).} and \textbf{Taste}\footnote{Defined in Visuddhimagga XIV.57 (see Footnote 2 for link).}) are objects of the other four senses. The objects of the tactile sense are Great Elements; they have a strong impact and therefore the accompanying \textbf{Feeling} of body-consciousness is either painful or pleasurable.\footnote{Mind Moment \textbf{17} is body-consciousness with painful \textbf{Feeling} and Mind Moment \textbf{24} is body-consciousness with pleasurable \textbf{Feeling}. Mind Moments \textbf{13}--\textbf{16} and \textbf{20}--\textbf{23} arise with indifferent \textbf{Feeling}.} The other sense objects have a weak impact, and therefore eye-consciousness, ear-consciousness, etc., are accompanied by indifferent \textbf{Feeling}.

Let’s look at \textbf{Visible-form} in a bit more detail. The Suttas and the Abhidhamma use the word \textit{rūpa} to mean both \textbf{Visible-form} and matter in general, so you must consider the context when you read the word “\textit{rūpa}.” \textbf{Visible-form} is a small patch of colour; this small patch of colour is what is actually seen by the eye.\footnote{The eye actually sees blue, green and red: \url{http://en.wikipedia.org/wiki/Photoreceptor_cell}} Once colour is seen by the eye, thinking takes place to create shapes and things. The eye does not see a cup; it sees a small patch of colour, and the mind then builds a larger image and creates a shape from colours; it creates things such as a cup. It can be difficult to know what \textbf{Visible-form} is because we are usually caught up in concepts such as a cup. The ear does not hear a song, the ear hears a single \textbf{Sound} and the mind uses many single \textbf{Sounds} to create a song.

\pagebreak

Using conventional language we might say, “I see my wife walking,” but the only Ultimate Reality in this example is a very small patch of colour somewhere in the field of vision.\footnote{Extend your arm and look at your thumbnail; this is the size of the portion of the field of vision that the retina can process at one time. The eye is constantly moving to build a complete visual image.} Everything else is added by the mind. As mentioned in the lesson on Mind Moments, whenever something is added by the mind, there is an opportunity for latent fears and latent \textbf{Attachment} to distort and misinterpret. Modern research into eye movement\footnote{\url{http://en.wikipedia.org/wiki/Eye_movement}} has shown that the mind filters the visual scene by causing the eye to focus on areas of high contrast, to focus on faces or on movement. In other words, the mind naturally distorts and filters what is perceived.

The same principle applies to other senses; for example, consider the difference in hearing a language you understand and hearing a language you do not understand. When hearing a language you understand, the mind automatically creates words and ideas to accompany the sounds.

Imagine that you are sitting quietly in a meditation hall and somebody starts speaking outside. If the mind is interested in what is being said, the senses are not being guarded. If the mind is angry at the speaker, the senses are not being guarded. Getting caught up in a sense object, looking for reasons not to be mindful of what is happening in the present moment, is not wholesome. If there is awareness of \textbf{Sound}, but nothing more, this is a wholesome Mind Moment and the senses are being guarded. Similarly, treating the \textbf{Hardness} of a diamond as being the same as the \textbf{Hardness} of a pebble is guarding the senses and not getting caught up in the concept of money.

Each sense object is either intrinsically undesirable or intrinsically desirable.\footnote{According to the Commentary to the Abhidhamma (\textit{Sammohavinodanī} 1, 10--11), an object’s intrinsic nature is according to “average” men such as accountants, government officials, land-owners and merchants. A masochist may enjoy pain, but pain is still intrinsically undesirable.} For example, the \textbf{Odour} of garbage is intrinsically undesirable while the \textbf{Odour} of coffee is intrinsically desirable. When the \textbf{Odour} of garbage enters the nostrils, this is a condition for Mind Moment \textbf{15} to arise; nose-consciousness that is a result of past unwholesome kamma. When the \textbf{Odour} of coffee enters the nostrils, it is a condition for Mind Moment \textbf{22} to arise; nose-consciousness that is a result of past wholesome kamma. Touching something that is very hot or very cold is a condition for Mind Moment \textbf{17} to arise; the result of past unwholesome kamma. Touching something that is close to body temperature is a condition for Mind Moment \textbf{24} to arise; the result of past wholesome kamma. This will be explored in more detail during the lesson on processes.

\subsection*{Sense Organs}

\begin{figure} [H]
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}
\noindent\begin{tabular}{P{0.25\textwidth} | C{.185\textwidth} C{0.185\textwidth} C{0.185\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Eye-sensitivity}\newline \textit{Cakkhu-pasāda} & Sensitivity to \textbf{Visible-forms} & Finding object among \textbf{Visible-forms} & Footing of eye-consciousness & 4 Great Elements, kamma-born desire to see &\\[9mm]
\textbf{Ear-sensitivity}\newline \textit{Sota-pasāda} & Sensitivity to \textbf{Sounds} & Finding object among \textbf{Sounds} & Footing of ear-consciousness & 4 Great Elements, kamma-born desire to hear &\\[9mm]
\textbf{Nose-sensitivity}\newline \textit{Ghāna-pasāda} & Sensitivity to \textbf{Odours} & Finding object among \textbf{Odours} & Footing of nose-consciousness & 4 Great Elements, kamma-born desire to smell &\\[9mm]
\textbf{Tongue-sensitivity}\newline \textit{Jivhā-pasāda} & Sensitivity to \textbf{Tastes} & Finding object among \textbf{Tastes} & Footing of tongue-consciousness & 4 Great Elements, kamma-born desire to taste &\\[9mm]
\textbf{Body-sensitivity}\newline \textit{Kāya-pasāda} & Sensitivity to tactile objects & Finding object among\newline tactile objects & Footing of body-consciousness & 4 Great Elements, kamma-born desire to feel &\\[9mm]
\bottomrule
\end{tabular}
\caption[]{The five sense organs.\footnotemark}
\end{figure}

\footnotetext{Explained in Chapter 3 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link).}

The next set of \textit{rūpas} are the sense organs. \textbf{Eye-sensitivity}\footnote{Defined in Visuddhimagga XIV.47 (see Footnote 2 for link).} is the portion of the eye that is capable of seeing colour; in other words, the retina. As mentioned in a previous lesson, the Suttas explain that when \textbf{Visible-form} co-exists with the sensitive part of the eye, then eye-consciousness arises naturally. This does not mean that eye-consciousness is created by \textbf{Visible-form} striking the sensitive part of the eye;\footnote{\textbf{Visible-form} is external, so it cannot physically strike the sensitive part of the eye.} it means that \textbf{Visible-form} and a working sensitive part of the eye are necessary conditions for eye-consciousness to arise.

Just as \textbf{Eye-sensitivity} is located in the eye, \textbf{Ear-sensitivity}\footnote{Defined in Visuddhimagga XIV.49 (see Footnote 2 for link).} is located in the ear, \textbf{Nose-sensitivity}\footnote{Defined in Visuddhimagga XIV.50 (see Footnote 2 for link).} is located in the nose, \textbf{Tongue-sensitivity}\footnote{Defined in Visuddhimagga XIV.51 (see Footnote 2 for link).} is located on the tongue, and \textbf{Body-sensitivity}\footnote{Defined in Visuddhimagga XIV.52 (see Footnote 2 for link).} is located all over the body except for the hair, teeth, and nails. \textbf{Body-sensitivity} is located wherever there are nerves to detect earth-element such as \textbf{Hardness}, fire-element such as \textbf{Heat} and air-element such as \textbf{Pressure}.

\pagebreak

The Abhidhamma also refers to the sense organs as doors. For example, \textbf{Eye-sensitivity} is the door through which the eye-consciousness Mind Moment interacts with the \textbf{Visible-form}. Seeing happens at the eye-door, hearing happens at the ear-door, and so on. The Abhidhamma extends this analogy and refers to thinking as happening at the mind-door. So there are six doors; five doors for the physical senses and the mind-door.

The Abhidhamma also refers to the sense organs as bases. A base supports the associated consciousness and is the location where the associated consciousness arises. The two types of eye-consciousness, Mind Moment \textbf{13} and Mind Moment \textbf{20}, arise at the eye-base, the retina. Similarly, the two types of ear-consciousness arise at the ear-base, the two types of nose-consciousness arise at the nose-base, the two types of tongue-consciousness arise at the tongue base and the two types of body-consciousness arise at the body-base.

Now we know where 10 of these Mind Moments, the two sets of fivefold sense consciousness, arise. But what about the remaining 79 Mind Moments that are listed in Appendix 2? The Abhidhamma mentions that these Mind Moments arise based on “that \textit{rūpa} upon which the mind depends,” and the Commentary calls it the “\textbf{Heart-base}.” Just as eye-base supports eye-consciousness, ear-base supports ear-consciousness and so on, the \textbf{Heart-base} supports all the Mind Moments except for the 10 sense-consciousness.

\subsection*{Heart-base}

\begin{figure} [H]
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}
\noindent\begin{tabular}{P{0.25\textwidth} | C{.185\textwidth} C{0.185\textwidth} C{0.185\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Heart-base}\newline \textit{Hadaya-vatthu} & Support for\newline the mind & Supporting the mind & Carrying the mind & Four Great Elements &\\[9mm]
\bottomrule
\end{tabular} 
\caption[]{The \textbf{Heart-base} \textit{rūpa}.\footnotemark}
\end{figure}

\footnotetext{Defined in Visuddhimagga XIV.60, explained in Chapter 5 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link).}

According to the Abhidhamma, Mind Moments start to arise in a new existence immediately after conception. At that time, there is no heart (and no brain) to support these Mind Moments. This is why the Abhidhamma does not specify an organ as the base for the mind but rather uses the description, “that \textit{rūpa} upon which the mind depends.”\footnote{The term “Heart-base” is from the Commentary. The \textit{Dhammasaṅgaṇī} omits “Heart-base” from the list of \textit{rūpas} and the reference to “that \textit{rūpa} upon which the mind depends” is from the \textit{Paṭṭhāna}.}

In ancient times, people recognized that information sensed by different parts of the body needed to be collected in the mind. The only thing that was seen moving around the body was blood and the heart plays an obvious role in pumping blood. This is the basis of the view that the heart is the base for the mind that continues today in common speech. For example, you don’t say, “I love you with all my brain!” Interestingly, there are cases of heart transplant recipients “inheriting” characteristics from the donor.\footnote{\url{http://www.namahjournal.com/doc/Actual/Memory-transference-in-organ-transplant-recipients-vol-19-iss-1.html}}

\subsection*{Nutriment}

\begin{figure} [H]
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}
\noindent\begin{tabular}{P{0.25\textwidth} | C{.185\textwidth} C{0.185\textwidth} C{0.185\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Nutriment}\newline \textit{Āhāra}/\textit{Ojā} & Nutritive\newline essence & Feeding kinds of matter & Consolidating & Physical bases that must be fed &\\[9mm]
\bottomrule
\end{tabular}
\caption[]{The \textbf{Nutriment} rūpa.\footnotemark}
\end{figure}

\footnotetext{Defined in Visuddhimagga XIV.70, explained in Chapter 2 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link).}

All groups of \textit{rūpas} include \textbf{Nutriment}; what we call “food” has a greater intensity of \textbf{Nutriment} \textit{rūpa}. \textbf{Nutriment} \textit{rūpa} can be ingested into the stomach or can come through an external application such as eye drops or skin cream. According to the Suttas,\footnote{SN 12.63: \url{http://www.accesstoinsight.org/tipitaka/sn/sn12/sn12.063.than.html}} just as the body depends on the \textbf{Nutriment} \textit{rūpa}, the mind depends on the nutriment of \textbf{Contact}, \textbf{Volition}, and consciousness.\footnote{In MN 9 (\url{http://www.accesstoinsight.org/tipitaka/mn/mn.009.than.html\#nutriments4}),\\ Sāriputta explains that properly discerning the four types of nutriment is Right View.}

\subsection*{Faculties}

\begin{figure} [H]
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}
\noindent\begin{tabular}{P{0.25\textwidth} | C{.185\textwidth} C{0.185\textwidth} C{0.185\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Femininity}\newline \textit{Itthindriya} & The state of\newline being female & Showing\newline “This is female” & Femininity in feature, mark, deportment & Four Great Elements &\\[9mm]
\textbf{Masculinity}\newline \textit{Purisindriya} & The state of\newline being male & Showing\newline “This is male” & Masculinity in feature, mark, deportment & Four Great Elements&\\[9mm]
\textbf{Life}\newline \textit{Jīvitindriya} & Maintaining conascent\newline matter & Making conascent\newline matter occur & Establishing presence of conascent matter & Four Great Elements &\\[9mm]
\bottomrule
\end{tabular}
\caption[]{The faculties.\footnotemark}
\end{figure}

\footnotetext{Explained in Chapter 5 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link).}

\textbf{Femininity-faculty}, \textbf{Masculinity-faculty}\footnote{Both defined in Visuddhimagga XIV.58 (see Footnote 2 for link).}, and \textbf{Life-faculty},\footnote{Defined in Visuddhimagga XIV.59 (see Footnote 2 for link).} are not localized; they are spread across the entire body just like the sense-organ of touch. Many women try to accentuate their \textbf{Femininity-faculty} and many men try to accentuate their \textbf{Masculinity-faculty}. This is \textbf{Attachment}; a desire to look feminine or to look masculine. 

\pagebreak

Reminds me of a joke: women get married hoping he will change and he doesn’t, men get married hoping she will not change and she does. Tendencies tend to stay and the body always changes. Not getting what you want is \textit{dukkha}.\footnote{According to SN 6.63 (\url{http://www.accesstoinsight.org/tipitaka/an/an06/an06.063.than.html\#part-6}), “Birth is stress, ageing is stress, death is stress; sorrow, lamentation, pain, distress, \& despair are stress; association with the unloved is stress; separation from the loved is stress; not getting what is wanted is stress. In short, the five clinging-aggregates are stress.”} When I was first married, I told my wife that I found her long hair to be very attractive. Twenty-five years later, my heart was filled with joy when she shaved her head to ordain temporarily as a Sayalay.

\subsection*{Material-\textit{rūpas} and Nominal-\textit{rūpas}}

The \textit{rūpas} discussed so far are “material-\textit{rūpas};” they have the three characteristics of \textit{anicca}, \textit{dukkha} and \textit{anattā}. The remaining \textit{rūpas} are “nominal-\textit{rūpas};” they are attributes or characteristics of the associated groups of material-\textit{rūpas} and do not have the three characteristics of \textit{anicca}, \textit{dukkha} and \textit{anattā}.

\subsection*{Space}

\begin{figure} [H]
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}
\noindent\begin{tabular}{P{0.25\textwidth} | C{.185\textwidth} C{0.185\textwidth} C{0.185\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Space}\newline \textit{Ākāsa} & Delimiting matter & Displaying the boundaries of matter & Non-touching\newline of matter & The matter delimited &\\[9mm]

\bottomrule
\end{tabular}
\caption[]{The \textbf{Space} rūpa.\footnotemark}
\end{figure}

\footnotetext{Defined in Visuddhimagga XIV.63, explained in Chapter 7 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link).}

The first nominal-\textit{rūpa} is \textbf{Space}. \textbf{Space} does not exist as an Ultimate Reality, but some meditation teachers\footnote{For example, in the Pa-Auk method of meditation.} advise their students to concentrate on the \textbf{Space} element to discern or infer the separation between different groups. Delimiting and separating groups allows the meditator to discern or infer the groups as being distinct. A rock is a cluster of many groups of \textit{rūpas}. Breaking up a rock shows there is \textbf{Space} between the individual groups. A group of \textit{rūpas} is infinitesimally small, conceptually similar to an atom. However, one should not think of individual \textit{rūpa} as being like subatomic particles. For example, a group of \textit{rūpas} will have both \textbf{Hardness} and \textbf{Heat}; the \textbf{Hardness} and \textbf{Heat} cannot be separated but they can be sensed individually. As an analogy, a person has both height and weight; height and weight cannot be separated but they can be measured individually.

\subsection*{Indications}

\begin{figure} [H]
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}
\noindent\begin{tabular}{P{0.25\textwidth} | C{.185\textwidth} C{0.185\textwidth} C{0.185\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Bodily-indication}\newline Bodily intimation\newline \textit{Kāya-viññatti} & Communicate & Displaying intention & Bodily excitement & Mind-born\newline \textbf{Supporting} &\\[12mm]
\textbf{Verbal-indication}\newline Verbal intimation\newline \textit{Vacī-viññatti} & Communicate & Displaying intention & Verbal excitement & Mind-born \textbf{Hardness} &\\[12mm]
\bottomrule
\end{tabular}
\caption[]{The two indications.\footnotemark}
\end{figure}

\footnotetext{Explained in Chapter 6 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link).}

The next two nominal-\textit{rūpas} are the indications; \textbf{Bodily-indication}\footnote{Defined in Visuddhimagga XIV.61 (see Footnote 2 for link).} and \textbf{Verbal-indication}.\footnote{Defined in Visuddhimagga XIV.62 (see Footnote 2 for link).} These are intentional means of self-expression that may communicate one’s thoughts to another person or even to an animal. Some Abhidhamma texts refer to these \textit{rūpas} as bodily intimation and verbal intimation.

Bodily movement involves a predominance of the mind-born air-element, a material-\textit{rūpa}. When the air-element of bodily movement is intentional, the group includes the nominal-\textit{rūpa} of \textbf{Bodily-indication}. \textbf{Bodily-indication} is an attribute of the bodily movement air-element. If I nod my head as I doze off, this may be the result of the mind-born air-element, perhaps neck muscles relaxing. If I nod my head to indicate that I agree with you, the group also includes the nominal-\textit{rūpa} of \textbf{Bodily-indication}. \textbf{Bodily-indication} is not a material-\textit{rūpa}; it is a special attribute of the mind-born air-element involved in bodily movement. \textbf{Bodily-indication} is not the gesture, it is the intentional aspect of a gesture that may communicate a message.

One’s voice arises from the change of the mind-born earth-element in the vocal cords. When the voice communicates a message, the group includes the nominal-\textit{rūpa} of \textbf{Verbal-indication}. \textbf{Verbal-indication} is an attribute of the mind-born earth-element that is the voice. If I use my voice to communicate a message to you, the group includes the nominal-\textit{rūpa} of \textbf{Verbal-indication}. \textbf{Verbal-indication} is not a material-\textit{rūpa}; it is a special attribute of the mind-born earth-element involved in speaking. \textbf{Verbal-indication} is not the \textbf{Sound}, it is the intentional aspect of making a \textbf{Sound} that communicates a message.

\textbf{Bodily-indication} and \textbf{Verbal-indication} arise with intention as special attributes of mind-born material-\textit{rūpas}. If my stomach rumbles, there is no intention involved so this is not a mind-born group. A rumbling stomach is a temperature-born \textbf{Sound} group.

\subsection*{Agility, Pliancy, and Adaptability}

\begin{figure} [H]
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}
\noindent\begin{tabular}{P{0.25\textwidth} | C{.185\textwidth} C{0.185\textwidth} C{0.185\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Agility}/\newline Lightness/Buoyancy\newline \textit{Rūpassa-lahutā} & Non-sluggishness & Removing heaviness & Quickness of change & Buoyant matter &\\[12mm]
\textbf{Pliancy}/\newline Elasticity/Malleability\newline \textit{Rūpassa-mudutā} & Non-rigidity & Removing\newline rigidity & Absence of opposition & Plastic matter &\\[12mm]
\textbf{Adaptability}/\newline Workability\newline \textit{Rūpassa-kammaññatā} & Workableness & Removing non-workableness & Non-weakness & Workable matter &\\[12mm]
\bottomrule
\end{tabular}
\caption[]{The \textbf{Agility}, \textbf{Pliancy}, and \textbf{Adaptability} rūpas.\footnotemark}
\end{figure}

\footnotetext{Defined in Visuddhimagga XIV.64, explained in Chapter 7 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link).}

The next three nominal-\textit{rūpas} are \textbf{Agility}, \textbf{Pliancy}, and \textbf{Adaptability}. These \textit{rūpas} arise in groups that are part of the body and are attributes of physical health. When they arise in a temperature-born group, the body is healthy because of suitable climate. When they arise in a mind-born group, the body is healthy because of faultless Mind Moments. When they arise in a nutriment-born group, the body is healthy because of nutritious food. \textbf{Agility}, \textbf{Pliancy}, and \textbf{Adaptability} are not distinct material-\textit{rūpas}; they are a sign that the body’s biological systems are in balance and healthy.

Buddhism emphasizes the importance of both mental health and physical health. Sometimes we need to be reminded about how important the mind is for the health of the body. In one Sutta, the Buddha visited a monk who was very ill.\footnote{SN 46.14: \url{http://www.accesstoinsight.org/tipitaka/sn/sn46/sn46.014.than.html}} The Buddha gave a short Dhamma talk and by reflecting on the Dhamma, the monk recovered from his sickness. The effect of the mind on the health of the body is well known to doctors. Mental stress leads to many health problems. A placebo will often improve a patient’s medical condition. We all know somebody whose positive attitude helped them through an illness. We should not wait until we are sick; wholesome Mind Moments will improve our health each moment and as it says in the Dhammapada, “Health is the greatest gift.”\footnote{See \url{http://www.tipitaka.net/tipitaka/dhp/verseload.php?verse=204}; in the Commentary to this verse, the Buddha advises King Pasenadi that it is not healthy to overeat.}

\textbf{Agility}, \textbf{Pliancy}, and \textbf{Adaptability} are also names of Mental Factors that make wholesome Mind Moments non-sluggish, non-rigid and non-judgemental. The \textit{rūpas} of \textbf{Agility}, \textbf{Pliancy} and \textbf{Adaptability} imply a material-\textit{rūpa} that is healthy. When the body is not healthy, it is sluggish, rigid and not workable; it lacks the \textit{rūpas} of \textbf{Agility}, \textbf{Pliancy} and \textbf{Adaptability}.

\subsection*{Characteristics}

\begin{figure} [H]
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}
\noindent\begin{tabular}{P{0.25\textwidth} | C{.185\textwidth} C{0.185\textwidth} C{0.185\textwidth} C{0.195\textwidth} m{0mm}}
\toprule
 & \tableheader{Characteristic} & \tableheader{Function} & \tableheader{Manifestation} & \tableheader{Proximate Cause}\\
\midrule
\textbf{Production}\newline \textit{Upacaya} & Setting up & Making matter emerge initially & Launching & Grown matter &\\[9mm]
\textbf{Continuity}\newline \textit{Santati} & Continuous occurrence & Linking/Binding without a break & Unbroken\newline series & Matter bound up without a break &\\[9mm]
\textbf{Decay}\newline \textit{Jaratā} & Maturing/\newline Ripening & Leading toward termination & Loss of\newline newness & Matter that is maturing/ripening &\\[9mm]
\textbf{Impermanence}\newline \textit{Aniccatā} & Complete breaking up & Making matter subside & Destruction\newline and fall & Matter that is breaking up &\\[9mm]
\bottomrule
\end{tabular}
\caption[]{The four characteristics.\footnotemark}
\label{Characteristics}
\end{figure}

\footnotetext{Explained in Chapter 8 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link).}

The final four nominal-\textit{rūpas} are \textbf{Production},\footnote{Defined in Visuddhimagga XIV.67 (see Footnote 2 for link).} \textbf{Continuity},\footnote{Defined in Visuddhimagga XIV.67 (see Footnote 2 for link).} \textbf{Decay},\footnote{Defined in Visuddhimagga XIV.68 (see Footnote 2 for link).} and \textbf{Impermanence}.\footnote{Defined in Visuddhimagga XIV.69 (see Footnote 2 for link).} These four \textit{rūpas} apply only to the body. \textbf{Production} starts from the moment of conception and continues until a complete set of \textit{rūpas} are created (the sense organs arise some time after conception). \textbf{Continuity} takes over once a complete set of \textit{rūpas} are created, and continuously refreshes the body throughout its lifespan. \textbf{Decay} is the inevitable ageing process such as grey hair and wrinkles that affect the body, and \textbf{Impermanence} is the death of the body.

\subsection*{Groups of \textit{Rūpas}}

We have finished describing the 28 \textit{rūpas}, so now let’s focus on how these \textit{rūpas} arise in groups.\footnote{Explained in Chapter 9 of “The Buddhist Teaching on Physical Phenomena” (see Footnote 2 for link). In Pāḷi, these groups are called \textit{kalāpa}, a term that was introduced in the Commentaries.} There are four temperature-born groups, nine kamma-born groups, six mind-born groups and two nutriment-born groups.

\pagebreak

\subsubsection*{Temperature-born groups of \textit{rūpas}}

\begin{figure} [H]
\centering
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}

\noindent\begin{tabular}{R{.37\textwidth} |
p{.045\textwidth} 
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}}
\toprule
& \tablevsubheader{Pure octad group}
& \tablevsubheader{Healthy octad}
& \tablevsubheader{Sound}
& \tablevsubheader{Healthy sound}
\\
\midrule
\hpadright{Eight inseperable \textit{rūpas}} & \tm & \tm & \tm & \tm
\\
\hpadright{\textbf{Sound}} & & &\tm & \tm
\\
\hpadright{\textbf{Agility}, \textbf{Pliancy} \& \textbf{Adaptability}} & &\tm & & \tm
\\
\bottomrule
\end{tabular}
\begin{center}
\tm\hspace{2mm} \textit{Rūpa} included in group
\end{center}
\caption{\textit{Rūpas} in the temperature-born groups.}
\end{figure}

The body is made up of temperature-born groups, kamma-born groups, mind-born groups and nutrition-born groups. Things external to the body are made up of either temperature-born pure-octad groups or temperature-born \textbf{Sound} groups. A cup of coffee does not make a \textbf{Sound}, so it is made up of temperature-born pure-octad groups. A fan makes a \textbf{Sound}, so it is made up of temperature-born \textbf{Sound} groups.

From a Buddhist perspective, a cup of coffee and a fan are equivalent; they have the ability to be taken as sense objects. Plants do not have consciousness or kamma and are also made up of temperature-born pure-octad groups, unless they are making a \textbf{Sound}.

“Octad” means a collection of eight things; in this case, the eight \textit{rūpas} of earth-element, water-element, fire-element, air-element, \textbf{Visible-form}, \textbf{Odour}, \textbf{Taste} and \textbf{Nutriment}. All of the groups of \textit{rūpas} include these eight \textit{rūpas} as a foundation.\footnote{These are called “the eight inseparables” because they always arise as the foundation for every \textit{kalāpa}.}

When the temperature-born pure-octad group is part of a healthy body, the \textit{rūpas} of \textbf{Agility}, \textbf{Pliancy} and \textbf{Adaptability} are present and this is the second group of \textit{rūpas}.

The temperature-born \textbf{Sound} group can be external to the body, such as a fan, or internal to the body such as a rumbling stomach. When the temperature-born \textbf{Sound} group is part of a healthy body, the \textit{rūpas} of \textbf{Agility}, \textbf{Pliancy} and \textbf{Adaptability} are present and this is the fourth group of \textit{rūpas}.

Fire-element arises in all groups of \textit{rūpas}. Irrespective of which group it arises in, fire-element can be a condition for the arising of a new temperature-born group of \textit{rūpas}. A rock continues to be a rock, moment after moment, because temperature-born groups of \textit{rūpas} arise repeatedly.

\subsubsection*{Kamma-born groups of \textit{rūpas}}

\begin{figure} [H]
\centering
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}

\noindent\begin{tabular}{R{.37\textwidth} |
p{.045\textwidth} 
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}}
\toprule
& \tablevsubheader{Vital nonad group}
& \tablevsubheader{Femininity}
& \tablevsubheader{Masculinity}
& \tablevsubheader{Heart-base}
& \tablevsubheader{Eye-senstivity}
& \tablevsubheader{Ear-senstivity}
& \tablevsubheader{Nose-senstivity}
& \tablevsubheader{Tongue-senstivity}
& \tablevsubheader{Body-sensitivity}
\\
\midrule
\hpadright{Eight inseperable \textit{rūpas}} & \tm & \tm & \tm & \tm & \tm & \tm & \tm & \tm & \tm
\\
\hpadright{\textbf{Eye-sensitivity}} & & & & & \tm & & & & 
\\
\hpadright{\textbf{Ear-sensiitivity}} & & & & & &\tm & & &
\\
\hpadright{\textbf{Nose-sensitivity}} & & & & & & & \tm & & 
\\
\hpadright{\textbf{Tongue-sensitivity}} & & & & & & & & \tm &
\\
\hpadright{\textbf{Body-sensitivity}} & & & & & & & & & \tm
\\
\hpadright{\textbf{Heart-base}} & & & &\tm & & & & &
\\
\hpadright{\textbf{Femininity}} & & \tm & & & & & & & 
\\
\hpadright{\textbf{Masculinity}} & & & \tm & & & & & &
\\
\hpadright{\textbf{Life}} & \tm & \tm & \tm & \tm & \tm & \tm &\tm & \tm & \tm
\\

\bottomrule
\end{tabular}
\begin{center}
\tm\hspace{2mm} \textit{Rūpa} included in group
\end{center}
\caption{\textit{Rūpas} in the kamma-born groups.}
\end{figure}

The next nine groups of \textit{rūpas} are kamma-born. These groups of \textit{rūpas} are part of the body. Life, gender, \textbf{Heart-base} and each of the five sense organs are the result of the same rebirth-linking kamma that gave rise to this existence.

In the past existence, there was \textbf{Attachment} to life, \textbf{Attachment} to femininity or masculinity and \textbf{Attachment} to sensing and thinking that led to this existence. An Arahat has uprooted all of these \textbf{Attachments} and therefore has no conditions for rebirth.

\subsubsection*{Mind-born groups of \textit{rūpas}}

\begin{figure} [H]
\centering
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}

\noindent\begin{tabular}{R{.37\textwidth} |
p{.045\textwidth} 
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}
p{.045\textwidth}}
\toprule
& \tablevsubheader{Pure octad group}
& \tablevsubheader{Healthy octad}
& \tablevsubheader{Bodily-indication}
& \tablevsubheader{Healthy bodily-indication}
& \tablevsubheader{Verbal-indication}
& \tablevsubheader{Healthy verbal-indication}
\\
\midrule
\hpadright{Eight inseperable \textit{rūpas}} & \tm & \tm & \tm & \tm & \tm & \tm
\\
\hpadright{\textbf{Sound}} & & & & &\tm & \tm
\\
\hpadright{\textbf{Bodily-indication}} & & &\tm & \tm & &
\\
\hpadright{\textbf{Verbal-indication}} & & & & &\tm & \tm
\\
\hpadright{\textbf{Agility}, \textbf{Pliancy} \& \textbf{Adaptability}} & & \tm & & \tm & & \tm
\\
\bottomrule
\end{tabular}
\begin{center}
\tm\hspace{2mm} \textit{Rūpa} included in group
\end{center}
\caption{\textit{Rūpas} in the mind-born groups.}
\end{figure}

The next six groups of \textit{rūpas} are mind-born. These groups of \textit{rūpas} are also part of the body. Most of the 89 Mind Moments create mind-born groups of \textit{rūpas}. The exceptions are the 10 sense-consciousness Mind Moments (they are too weak), and the four Life-continuum Mind Moments in the Immaterial Sphere. I will explain more about these Mind Moments in the next lesson when I discuss Realms of Existence.

\begin{figure}[H]
\centering

\begin{center}
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}

\noindent\begin{tabular}{L{.29\textwidth}P{.1\textwidth}|p{.04\textwidth}p{.04\textwidth}p{.04\textwidth}p{.04\textwidth}}
\toprule
& & \tablevsubheaderhacksmall{Can create a new mind-born group} & \tablevsubheaderhacksmall{Can cause a change of posture} & \tablevsubheaderhacksmall{Can cause bodily \& verbal indications} & \tablevsubheaderhacksmall{Can cause smiling} \\
\midrule
\multirow{2}{.29\textwidth}{Danger Zone} & \textbf{1}--\textbf{4} & \tmsmall & \tmsmall & \tmsmall & \tmsmall \\
& \textbf{5}--\textbf{12} & \tmsmall & \tmsmall & \tmsmall & \\
\midrule
\multirow{6}{.29\textwidth}{Sensing Zone} & \textbf{13}--\textbf{17} & & & & \\
& \textbf{18}--\textbf{19} & \tmsmall & & & \\
& \textbf{20}--\textbf{24} & & & & \\
& \textbf{25}--\textbf{28} & \tmsmall & & & \\
& \textbf{29} & \tmsmall & \tmsmall & \tmsmall & \\
& \textbf{30} & \tmsmall & \tmsmall & \tmsmall & \tmsmall \\
\midrule
\multirow{2}{.29\textwidth}{Faultless Zone} & \textbf{31}--\textbf{34} & \tmsmall & \tmsmall & \tmsmall & \tmsmall \\
& \textbf{35}--\textbf{38} & \tmsmall & \tmsmall & \tmsmall & \\
\midrule
Sense Sphere Resultant & \textbf{39}--\textbf{46} & \tmsmall & & & \\
\midrule
\multirow{2}{.29\textwidth}{Sense Sphere Functional} & \textbf{47}--\textbf{50} & \tmsmall & \tmsmall & \tmsmall & \tmsmall \\
& \textbf{50}--\textbf{54} & \tmsmall & \tmsmall & \tmsmall & \\
\midrule
\multirow{3}{.29\textwidth}{Fine Material Sphere} & \textbf{55}--\textbf{59} & \tmsmall & \tmsmall & & \\
& \textbf{60}--\textbf{64} & \tmsmall & & & \\
& \textbf{65}--\textbf{69} & \tmsmall & \tmsmall & & \\
\midrule
\multirow{3}{.29\textwidth}{Immaterial Sphere} & \textbf{70}--\textbf{73} & \tmsmall & \tmsmall & & \\
& \textbf{74}--\textbf{77} & & & & \\
& \textbf{78}--\textbf{81} & \tmsmall & \tmsmall & & \\
\bottomrule
\end{tabular}
\end{center}

\begin{center}
\noindent
\tmsmall \hspace{2mm} Mind Moment can cause this kind of mind-born group

\end{center}

\caption{Mind Moments and the mind-born groups that they can create.}
\label{fig:MindBorn}
\end{figure}

\pagebreak

\subsubsection*{Nutriment-born groups of \textit{rūpas}}

\begin{figure}[H]
\centering
\setlength{\tabcolsep}{0pt}
\renewcommand{\arraystretch}{1.1}

\noindent\begin{tabular}{R{.37\textwidth} |
p{.045\textwidth} 
p{.045\textwidth}}
\toprule
& \tablevsubheader{Pure octad group}
& \tablevsubheader{Healthy octad}
\\
\midrule
\hpadright{Eight inseperable \textit{rūpas}} & \tm & \tm
\\
\hpadright{\textbf{Agility}, \textbf{Pliancy} \& \textbf{Adaptability}} & &\tm
\\
\bottomrule
\end{tabular}
\begin{center}
\tm\hspace{2mm} \textit{Rūpa} included in group
\end{center}
\caption{\textit{Rūpas} in the nutriment-born groups.}
\end{figure}

The last two groups of \textit{rūpas} are nutriment-born. Nutriment-born groups of \textit{rūpas} are also part of the body. According to the Commentary, Nutriment \textit{rūpa} has the ability to produce new, nutriment-born groups of \textit{rūpas} for up to seven days after ingestion.

Most parts of the body include six groups of \textit{rūpas}: temperature-born pure-octad group (with or without health), kamma-born vital-nonad group, kamma-born gender (either femininity or masculinity) group, kamma-born Body-sensitivity group, mind-born pure-octad group (with or without health) and nutriment-born pure-octad group (with or without health). In other words, most parts of the body include three pure-octad groups (with or without health), vital-nonad group, either Femininity or Masculinity and Body-sensitivity. The physical eye includes seven groups of \textit{rūpas}; the same six groups of \textit{rūpas} as the rest of the body plus the kamma-born \textbf{Eye-sensitivity} group.

\subsection*{Linkage to \textit{Satipaṭṭhāna} Sutta}

Please refer to the \textit{Satipaṭṭhāna} Sutta in Appendix 1. “The contemplation of the body” section, paragraphs 4 through 31 has many exercises dealing with \textit{rūpa}. Different meditation teachers tend to focus on different aspects of contemplation of the body, but most of them involve \textit{rūpa}. To conclude the section of the lesson on \textit{rūpa}, I will give a couple of examples.

The most obvious example is paragraph 17, where the meditator is asked to consider the body as consisting of the four Great Elements.\footnote{See Visuddhimagga XI27--30 (See Footnote 2 for link).} Focusing on the four Great Elements is an exercise in depersonalizing the body. The analogy in paragraph 18 of a butcher slaughtering a cow and dividing it into portions is explained in the Commentary as follows: when it is feeding in the pasture, being led to the slaughterhouse and when it is being killed, there is a perception of “cow.” When it is divided into portions, the perception of “cow” is replaced with a perception of “meat,” “blood” and “bones.” Considering the “cow” as its constituent parts depersonalizes the “cow.”\footnote{According to the Commentary, “the junction of four high roads” refers to the four postures of going, standing, sitting and lying down; in other words, at all times.}

According to the Commentary, directly contemplating the body as being composed of the four Great Elements is suitable for those with quick understanding.\footnote{See Visuddhimagga IX.28 (see Footnote 2 for link).} According to the Commentary, those who do not have quick understanding should reflect on the four Great Elements by way of the parts of the body.\footnote{See Visuddhimagga IX.31 (see Footnote 2 for link).} 

As explained in the Suttas, the earth-element can be contemplated upon by reflecting on the hair of the head, hair of the body, nails, teeth, skin, flesh, sinews, bones, marrow, kidney, heart, liver, midriff, spleen, lungs, intestines, mesentery, gorge, and feces. The water-element can be contemplated by reflecting on bile, phlegm, pus, blood, sweat, fat, tears, grease, saliva, nasal mucus, synovial fluid, and urine. The fire-element can be contemplated by reflecting on that by which the body is warmed, aged, and consumed with fever; and that by which what is eaten, drunk, chewed, and savoured gets properly digested. Finally, the air-element can be contemplated by reflecting on up-going winds (such as burps), down-going winds (such as passing gas), winds in the stomach, winds in the intestines, winds that course through the body, and in-and-out breathing.\footnote{MN 28: \url{http://www.accesstoinsight.org/tipitaka/mn/mn.028.than.html}}

\pagebreak

\subsection*{Summary of Key Points}

\begin{itemize}

\item \textit{Rūpas} include all non-mental phenomena; a definition that is wider than the conventional translation of “matter.”

\item Buddhism focuses on spiritual development, so the Buddhist analysis of \textit{rūpa} focuses on sense objects, the sense organs, and the body. 

\begin{itemize}

\item Three of the “Four Great Elements” (\textbf{Hardness} or earth-element, \textbf{Heat} or fire-element, and \textbf{Supporting} or air-element) are objects of the tactile sense. The fourth Great Element (\textbf{Cohesion} or water-element) is experienced by the mind.

\item The sense objects of \textbf{Visible-form}, \textbf{Sound}, \textbf{Odour} and \textbf{Taste} are the sense objects of the remaining senses.

\item The five sense organs (\textbf{Eye-sensitivity}, \textbf{Ear-sensitivity}, \textbf{Nose-sensitivity}, \textbf{Tongue-sensitivity} and \textbf{Body-sensitivity}) are able to detect sense objects (for example, \textbf{Eye-sensitivity} is the portion of the retina that responds to light).

\item The \textbf{Heart-base} supports all Mind Moments except the sense-consciousness Mind Moments; the sense-consciousness Mind Moments are supported by their respective sense organ (for example, the eye-consciousness Mind Moment arises at the retina).

\item \textbf{Nutriment} \textit{rūpa} includes food (though all groups of \textit{rūpas} include nutriment).

\item The three faculties (\textbf{femininity}, \textbf{masculinity}, and \textbf{life}).

\item The remaining \textit{rūpas} are nominal; they are attributes of the associated group (nominal \textit{rūpas} do not have the characteristics of \textit{anicca}, \textit{dukkha} and \textit{anattā}):

\begin{itemize}

\item \textbf{Space} separates the groups of \textit{rūpas}.

\item \textbf{Bodily-indication} and \textbf{verbal-indication} communicate information.

\item \textbf{Agility}, \textbf{pliancy}, and \textbf{adaptability} are the attributes of physical health.

\item The four characteristics (\textbf{production}, \textbf{continuity}, \textbf{decay}, and \textbf{impermanence}) represent the stages of life.

\end{itemize}

\end{itemize}

\item \textit{Rūpas} arise in groups:

\begin{itemize}

\item Non-living things are made up of \textbf{temperature-born pure-octad groups} or \textbf{temperature-born \textbf{Sound} groups}. Their relevance in the Buddhist analysis of \textit{rūpa} is that non-living things are merely sense objects.

\item Living things are made of all groups of \textit{rūpas} (four temperature-born groups, nine kamma-born groups, six mind-born groups and two nutriment-born groups).

\end{itemize}

\item The \textbf{\textit{R}} \textbf{\textit{A}} \textbf{\textit{D}} \textbf{\textit{I}} \textbf{\textit{CA}} \textbf{\textit{L}} process (\textbf{\textit{R}}ecognize, \textbf{\textit{A}}ccept, \textbf{\textit{D}}epersonalize, \textbf{\textit{I}}nvestigate, \textbf{\textit{C}}ontemplate \textbf{\textit{A}}\textit{nicca}/\textbf{\textit{C}}ontemplate \textbf{\textit{A}}\textit{nattā}, \textbf{\textit{L}}et go) can also be applied to \textit{rūpa} (except for “nominal” \textit{rūpas}).

\end{itemize}

Finally, in my opinion, the most important thing to remember about \textit{rūpa} is that, in the context of spiritual development, its main relevance is as a sense object. We don’t touch a pebble; \textbf{Hardness}, \textbf{Heat}, and \textbf{Pressure} are experienced. “Pebble” is a concept. We don’t see a car. \textbf{Visible-form} is a small patch of colour experienced by the eye. “Car” is a concept. Guarding the senses means not getting caught up in concepts.

\newpage

\subsection*{Questions \& Answers}

\question{According to Buddhism, what is the role of the brain?}

In most of the Suttas (including paragraph 15 of the \textit{Satipaṭṭhāna} Sutta), there is a list of 31 parts of the body used to contemplate the repulsiveness of the body. Some Suttas (added at a much later date) expand this list to 32 parts of the body by adding the brain.\footnote{Khp 3: \url{http://www.accesstoinsight.org/tipitaka/kn/khp/khp.1-9x.piya.html\#khp-3}} Visuddhimagga VIII.126 describes the brain as the marrow of the skull bone and gives it no special significance (nothing related to the mind). In fact, Visuddhimagga VIII.136 describes snot coming out of the nose as originating in the brain!\footnote{I am glad that my doctor did not get his medical knowledge from the Visuddhimagga! ☺} The Suttas and the Abhidhamma associate sense-consciousness with the sense organs but do not associate the mind with any organ (see discussion on the \textbf{Heart-base} \textit{rūpa}).

In an article, “Is your brain really necessary?”\footnote{\url{http://www.rifters.com/real/articles/Science\_No-Brain.pdf}} Dr. John Lorber says, “There’s a young student at this university, who has an IQ of 126, has gained a first-class honours degree in mathematics, and is socially completely normal. And yet the boy has virtually no brain. When we did a brain scan on him, we saw that instead of the normal 4.5-centimeter thickness of brain tissue between the ventricles and the cortical surface, there was just a thin layer of mantle measuring a millimeter or so. His cranium is filled mainly with cerebrospinal fluid.” In my opinion, there are still many gaps in our scientific knowledge regarding the relationship between mind and body.

\question{The Suttas\footnote{SN 22.48: \url{http://www.accesstoinsight.org/tipitaka/sn/sn22/sn22.048.than.html}} describe \textit{rūpas} as “past, future, or present; internal or external; blatant or subtle; common or sublime.” What does this mean?}

I believe that the intention of the Buddha was to indicate that all \textit{rūpas}, no matter how they might be subdivided or analyzed, are to be included. The Commentary to the Abhidhamma analyzes \textit{rūpa} according to these categories, but I treat this part of the Commentary as more of a scholarly exercise rather than being helpful, practical or useful.

\question{According to the Buddhist way of describing the body, how can organs from a dead person be successfully transplanted into a living person?}

An organ coming from a donor will have temperature-born groups of \textit{rūpas} and nutriment-born groups of \textit{rūpas}. Because the donor is dead, there are no kamma-born groups of \textit{rūpas} or mind-born groups of \textit{rūpas} in the organ. After the transplant, the organ will have all four types of groups of \textit{rūpas}. The new kamma-born groups of \textit{rūpas} and the new mind-born groups of \textit{rūpas} will come from the kamma and the mind of the organ recipient.

\pagebreak

\question{How does the Abhidhamma explain how plants respond to music?}

According to the Abhidhamma, plants do not have consciousness nor do they have kamma. Therefore, plants' response to music cannot involve ear-consciousness Mind Moments or \textbf{Ear-sensitivity} \textit{rūpa}; plants' response to music must involve some other mechanism and this is not described in the Abhidhamma.

When I am asked questions such as these, I try not to extend the Abhidhamma or the Buddha's teachings into areas for which they are not intended. I ask myself, ``Are these just theoretical questions? Do they have any relevance to spiritual development?"

\question{You mentioned that gender is determined by kamma. Is being reborn as a female the result of past unwholesome kamma?}

Nowhere in the Vinaya, in the Suttas or in the Abhidhamma does it imply that rebirth as a female is the result of past unwholesome kamma (though there are some stories in the Commentaries that reflect this sexist attitude).\footnote{For example, \url{http://www.tipitaka.net/tipitaka/dhp/verseload.php?verse=043}} Women were able to become Arahats, making them the spiritual equal of men. One of the books of the \textit{Tipiṭaka} is dedicated to verses written by nuns.\footnote{\url{http://en.wikipedia.org/wiki/Therigatha}} The Buddha did say that a Buddha cannot be a woman\footnote{MN 115: \url{http://awake.kiev.ua/dhamma/tipitaka/2Sutta-Pitaka/2Majjhima-Nikaya/Majjhima3/115-bahudhatuka-e.html}} (of course, a woman in this life could be reborn as a man in a future life and become a Buddha in that life). When King Pasenadi was displeased at having a daughter, the Buddha said, “A woman may turn out better than a man.”\footnote{SN 3.16 (The on-line translation is a bit convoluted, so I am using Bhikkhu Bodhi’s translation): \url{http://metta.lk/tipitaka/2Sutta-Pitaka/3Samyutta-Nikaya/Samyutta1/03-Kosala-Samyutta/02-Aputtakavaggo-e.html}}

Some monks object to the ordination of Theravāda Bhikkhunī. Their objection is not based on any perceived inferiority of females; it is based on their interpretation of what is allowable under the Vinaya. I have found the following website to be an excellent source of information: \url{http://sites.google.com/site/santipada/bhikkhunifaq} Another source of information is an article, “The Legality of Bhikkhunī Ordination” by Ven. Anālayo: \url{http://www.buddhismuskunde.uni-hamburg.de/pdf/5-personen/analayo/legality.pdf} I am a strong supporter of Theravāda Bhikkhunī ordination. 

There is a story in the Commentary (\url{http://www.tipitaka.net/tipitaka/dhp/verseload.php?verse=043}) about a layman with two sons. He sees an Arahat and thinks, “I wish that monk were my wife or that my wife had a beautiful complexion like that monk.” The layman immediately becomes a woman. Ashamed, she moves to another city where she marries and has children as a woman. Later, she meets the same Arahat and apologizes. After apologizing, she becomes a man again. I find this story objectionable for because it implies that being a woman is the result of unwholesome kamma.